---
title: "SLAMSeq 2021_20_koertel final analysis"
author:
- name: Cornelia Rueckle
  affiliation: c.rueckle@imb-mainz.de
output:
  BiocStyle::pdf_document
abstract: |
  Analysis of SLAM Seq data
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message=F, warning=F, fig.path = "./figures")
```


```{r package import, echo=FALSE, message=FALSE, warning=FALSE}
## load libraries
library(gridExtra)
library(stringr)# needed for str_sub()
library(ggplot2)
library(dplyr)
library(minpack.lm)
library(rtracklayer)



level_order_mouse <- c(paste("chr", seq(1:19), sep = ""), "chrX")
```

# Processing data with SLAMDUNK

Reads were mapped using SLAMDUNK with default parameters and: -5 12 -n 100 -m

Reads were filtered using SLAMDUNK filter with default parameters

For SNP annotation, all unlabelled samples were merged and SNPs were called using SLAMDUNK snp with default parameters and  -f 0.2.

Transition rates were calculated SLAMDUNK count with default parameters providing the SNP annotation file of unlabelled samples (-v).

If more than one UTR per gene was remaining, they were collapsed using SLAMDUNK collapse

# How many reads are mapped



# Data preparation

For each time-point, four replicate experiments have been conducted. For each time-point, I will generate one file which includes the mean T2C convertions from all replicates and also the mean T coverage (will be used for expression estimation). 

+ no4SU (DMSOx4, STM2457 x 4)
+ T0 (DMSOx4, STM2457 x 4)
+ T1 (DMSOx4, STM2457 x 4)
+ T2 (DMSOx4, STM2457 x 4)
+ T3 (DMSOx4, STM2457 x 4)
+ T4 (DMSOx4, STM2457 x __3__)
+ T5 (DMSOx4, STM2457 x 4)
+ T6 (DMSOx4, STM2457 x 4)
+ T7 (DMSOx4, STM2457 x 4)

*One replicate in STM is missing due to too low RNA (STM2457 T4)*

```{r Data import, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}


#Import data
opt = data.frame(filepath="~/Documents/SLAM_seq/imb_koenig_2021_20_koertel_SLAM_seq/All_FC/Analysis/collapse/")

##Import datasets
files   <- list.files(opt$filepath,pattern="*.csv",full.names=F)

S4U.data        <- lapply(paste0(opt$filepath,files),function(x) {read.csv(x, header = TRUE, sep="\t",  comment.char = "#")})

#Reduce names of files to only important names
names(S4U.data) <- gsub(".fastq.gz.merged.fastq_slamdunk_mapped_filtered_tcount_collapsed.csv", "", files)
names(S4U.data) <-gsub(".*imb_koenig_2021_20_","",names(S4U.data))

data_DMSO <- S4U.data[grep("DMSO", names(S4U.data))]
stopifnot(length(data_DMSO)==36)
data_STM2457 <- S4U.data[grep("STM", names(S4U.data))]
stopifnot(length(data_STM2457)==35)




```


# Incorporation rate
## Check conversion rates

```{r Conversion rates, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}


ploting_data <- S4U.data


for (i in 1:length(S4U.data)){
  S4U.data[[i]]$name <- names(S4U.data[i])}

S4U.data$name <-NULL

plot_data <-S4U.data %>%
  bind_rows() 

plot_data$Incoorp_rate_percent <- (plot_data$conversionsOnTs/plot_data$coverageOnTs)*100

#ggplot(data=plot_data)+
  #geom_boxplot(aes(x=name, y=Incoorp_rate_percent))

plot_data$name <-str_sub(plot_data$name, 4, -4)

mean(c(na.omit((plot_data[plot_data$name=="T0_DMSO",]$Incoorp_rate_percent),
       na.omit(plot_data[plot_data$name=="T0_STM2457",]$Incoorp_rate_percent))))

ggplot()+
  geom_boxplot(data=plot_data, aes(x=name, y=Incoorp_rate_percent, fill=name), outlier.shape = NA)+
    #scale_fill_brewer()+
  coord_cartesian(ylim = c(0,2))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Incoorporation rate [%]")+
  theme(legend.position='none')


```


# Make mean

In order to calculate half-lives, mean transition rates for each sample are from all replicates.
```{r Make mean incoorp, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}

######################################################
######################################################
######################################################

### DMSO
data_DMSO_no4SU <- S4U.data$`01_no4SU_DMSO_01`


data_DMSO_no4SU$conversionRate_mean=(as.numeric(S4U.data$`01_no4SU_DMSO_01`$conversionRate)+as.numeric(S4U.data$`02_no4SU_DMSO_02`$conversionRate)+as.numeric(S4U.data$`03_no4SU_DMSO_03`$conversionRate)+as.numeric(S4U.data$`04_no4SU_DMSO_04`$conversionRate))/4


data_DMSO_no4SU$coverageOnTs_mean=(as.numeric(S4U.data$`01_no4SU_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`02_no4SU_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`03_no4SU_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`04_no4SU_DMSO_04`$coverageOnTs))/4


### DMSO T0

data_DMSO_T0 <- S4U.data$`09_T0_DMSO_01`

data_DMSO_T0$conversionRate_mean=(as.numeric(S4U.data$`09_T0_DMSO_01`$conversionRate)+as.numeric(S4U.data$`10_T0_DMSO_02`$conversionRate)+as.numeric(S4U.data$`11_T0_DMSO_03`$conversionRate)+as.numeric(S4U.data$`12_T0_DMSO_04`$conversionRate))/4

data_DMSO_T0$coverageOnTs_mean=(as.numeric(S4U.data$`09_T0_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`10_T0_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`11_T0_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`12_T0_DMSO_04`$coverageOnTs))/4

### DMSO T1

data_DMSO_T1 <- S4U.data$`17_T1_DMSO_01`

data_DMSO_T1$conversionRate_mean=(as.numeric(S4U.data$`17_T1_DMSO_01`$conversionRate)+as.numeric(S4U.data$`18_T1_DMSO_02`$conversionRate)+as.numeric(S4U.data$`19_T1_DMSO_03`$conversionRate)+as.numeric(S4U.data$`20_T1_DMSO_04`$conversionRate))/4

data_DMSO_T1$coverageOnTs_mean=(as.numeric(S4U.data$`17_T1_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`18_T1_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`19_T1_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`20_T1_DMSO_04`$coverageOnTs))/4
### DMSO T2

data_DMSO_T2 <- S4U.data$`25_T2_DMSO_01`

data_DMSO_T2$conversionRate_mean=(as.numeric(S4U.data$`25_T2_DMSO_01`$conversionRate)+as.numeric(S4U.data$`26_T2_DMSO_02`$conversionRate)+as.numeric(S4U.data$`27_T2_DMSO_03`$conversionRate)+as.numeric(S4U.data$`28_T2_DMSO_04`$conversionRate))/4


data_DMSO_T2$coverageOnTs_mean=(as.numeric(S4U.data$`25_T2_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`26_T2_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`27_T2_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`28_T2_DMSO_04`$coverageOnTs))/4
### DMSO T3

data_DMSO_T3 <- S4U.data$`33_T3_DMSO_01`

data_DMSO_T3$conversionRate_mean=(as.numeric(S4U.data$`33_T3_DMSO_01`$conversionRate)+as.numeric(S4U.data$`34_T3_DMSO_02`$conversionRate)+as.numeric(S4U.data$`35_T3_DMSO_03`$conversionRate)+as.numeric(S4U.data$`36_T3_DMSO_04`$conversionRate))/4

data_DMSO_T3$coverageOnTs_mean=(as.numeric(S4U.data$`33_T3_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`34_T3_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`35_T3_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`36_T3_DMSO_04`$coverageOnTs))/4

### DMSO T4

data_DMSO_T4 <- S4U.data$`41_T4_DMSO_01`

data_DMSO_T4$conversionRate_mean=(as.numeric(S4U.data$`41_T4_DMSO_01`$conversionRate)+as.numeric(S4U.data$`42_T4_DMSO_02`$conversionRate)+as.numeric(S4U.data$`43_T4_DMSO_03`$conversionRate)+as.numeric(S4U.data$`44_T4_DMSO_04`$conversionRate))/4

data_DMSO_T4$coverageOnTs_mean=(as.numeric(S4U.data$`41_T4_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`42_T4_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`43_T4_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`44_T4_DMSO_04`$coverageOnTs))/4

### DMSO T5

data_DMSO_T5 <- S4U.data$`48_T5_DMSO_01`

data_DMSO_T5$conversionRate_mean=(as.numeric(S4U.data$`48_T5_DMSO_01`$conversionRate)+as.numeric(S4U.data$`49_T5_DMSO_02`$conversionRate)+as.numeric(S4U.data$`50_T5_DMSO_03`$conversionRate)+as.numeric(S4U.data$`51_T5_DMSO_04`$conversionRate))/4

data_DMSO_T5$coverageOnTs_mean=(as.numeric(S4U.data$`48_T5_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`49_T5_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`50_T5_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`51_T5_DMSO_04`$coverageOnTs))/4

### DMSO T6

data_DMSO_T6 <- S4U.data$`56_T6_DMSO_01`

data_DMSO_T6$conversionRate_mean=(as.numeric(S4U.data$`56_T6_DMSO_01`$conversionRate)+as.numeric(S4U.data$`57_T6_DMSO_02`$conversionRate)+as.numeric(S4U.data$`58_T6_DMSO_03`$conversionRate)+as.numeric(S4U.data$`59_T6_DMSO_04`$conversionRate))/4

data_DMSO_T6$coverageOnTs_mean=(as.numeric(S4U.data$`56_T6_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`57_T6_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`58_T6_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`59_T6_DMSO_04`$coverageOnTs))/4

### DMSO T6

data_DMSO_T7 <- S4U.data$`64_T7_DMSO_01`

data_DMSO_T7$conversionRate_mean=(as.numeric(S4U.data$`64_T7_DMSO_01`$conversionRate)+as.numeric(S4U.data$`65_T7_DMSO_02`$conversionRate)+as.numeric(S4U.data$`66_T7_DMSO_03`$conversionRate)+as.numeric(S4U.data$`67_T7_DMSO_04`$conversionRate))/4

data_DMSO_T7$coverageOnTs_mean=(as.numeric(S4U.data$`64_T7_DMSO_01`$coverageOnTs)+as.numeric(S4U.data$`65_T7_DMSO_02`$coverageOnTs)+as.numeric(S4U.data$`66_T7_DMSO_03`$coverageOnTs)+as.numeric(S4U.data$`67_T7_DMSO_04`$coverageOnTs))/4

#data_DMSO <- data_DMSO[grep("no4SU", names(data_DMSO))][grep("conversionRate", )]
#grep("conversionRate", colnames(data_DMSO[grep("no4SU", names(data_DMSO))]))

DMSO_4SU_mean_data <- list(data_DMSO_T0,
                           data_DMSO_T1,
                           data_DMSO_T2,
                           data_DMSO_T3,
                           data_DMSO_T4,
                           data_DMSO_T5,
                           data_DMSO_T6,
                           data_DMSO_T7)

names(DMSO_4SU_mean_data) <- c("DMSO_T0",
                           "DMSO_T1",
                           "DMSO_T2",
                           "DMSO_T3",
                           "DMSO_T4",
                           "DMSO_T5",
                           "DMSO_T6",
                           "DMSO_T7")

######################################################
######################################################
######################################################


### STM2457
data_STM2457_no4SU <- S4U.data$`05_no4SU_STM2457_01`


data_STM2457_no4SU$conversionRate_mean=(as.numeric(S4U.data$`05_no4SU_STM2457_01`$conversionRate)+as.numeric(S4U.data$`06_no4SU_STM2457_02`$conversionRate)+as.numeric(S4U.data$`07_no4SU_STM2457_03`$conversionRate)+as.numeric(S4U.data$`08_no4SU_STM2457_04`$conversionRate))/4

data_STM2457_no4SU$coverageOnTs_mean=(as.numeric(S4U.data$`05_no4SU_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`06_no4SU_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`07_no4SU_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`08_no4SU_STM2457_04`$coverageOnTs))/4

### STM2457 T0

data_STM2457_T0 <- S4U.data$`13_T0_STM2457_01`

data_STM2457_T0$conversionRate_mean=(as.numeric(S4U.data$`13_T0_STM2457_01`$conversionRate)+as.numeric(S4U.data$`14_T0_STM2457_02`$conversionRate)+as.numeric(S4U.data$`15_T0_STM2457_03`$conversionRate)+as.numeric(S4U.data$`16_T0_STM2457_04`$conversionRate))/4

data_STM2457_T0$coverageOnTs_mean=(as.numeric(S4U.data$`13_T0_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`14_T0_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`15_T0_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`16_T0_STM2457_04`$coverageOnTs))/4

### STM2457 T1

data_STM2457_T1 <- S4U.data$`21_T1_STM2457_01`

data_STM2457_T1$conversionRate_mean=(as.numeric(S4U.data$`21_T1_STM2457_01`$conversionRate)+as.numeric(S4U.data$`22_T1_STM2457_02`$conversionRate)+as.numeric(S4U.data$`23_T1_STM2457_03`$conversionRate)+as.numeric(S4U.data$`24_T1_STM2457_04`$conversionRate))/4


data_STM2457_T1$coverageOnTs_mean=(as.numeric(S4U.data$`21_T1_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`22_T1_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`23_T1_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`24_T1_STM2457_04`$coverageOnTs))/4

### STM2457 T2

data_STM2457_T2 <- S4U.data$`29_T2_STM2457_01`

data_STM2457_T2$conversionRate_mean=(as.numeric(S4U.data$`29_T2_STM2457_01`$conversionRate)+as.numeric(S4U.data$`30_T2_STM2457_02`$conversionRate)+as.numeric(S4U.data$`31_T2_STM2457_03`$conversionRate)+as.numeric(S4U.data$`32_T2_STM2457_04`$conversionRate))/4

data_STM2457_T2$coverageOnTs_mean=(as.numeric(S4U.data$`29_T2_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`30_T2_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`31_T2_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`32_T2_STM2457_04`$coverageOnTs))/4
### STM2457 T3

data_STM2457_T3 <- S4U.data$`37_T3_STM2457_01`

data_STM2457_T3$conversionRate_mean=(as.numeric(S4U.data$`37_T3_STM2457_01`$conversionRate)+as.numeric(S4U.data$`38_T3_STM2457_02`$conversionRate)+as.numeric(S4U.data$`39_T3_STM2457_03`$conversionRate)+as.numeric(S4U.data$`40_T3_STM2457_04`$conversionRate))/4

data_STM2457_T3$coverageOnTs_mean=(as.numeric(S4U.data$`37_T3_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`38_T3_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`39_T3_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`40_T3_STM2457_04`$coverageOnTs))/4

### STM2457 T4

data_STM2457_T4 <- S4U.data$`45_T4_STM2457_01`

data_STM2457_T4$conversionRate_mean=(as.numeric(S4U.data$`45_T4_STM2457_01`$conversionRate)+as.numeric(S4U.data$`46_T4_STM2457_02`$conversionRate)+as.numeric(S4U.data$`47_T4_STM2457_03`$conversionRate))/3


data_STM2457_T4$coverageOnTs_mean=(as.numeric(S4U.data$`45_T4_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`46_T4_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`47_T4_STM2457_03`$coverageOnTs))/3

### STM2457 T5

data_STM2457_T5 <- S4U.data$`52_T5_STM2457_01`

data_STM2457_T5$conversionRate_mean=(as.numeric(S4U.data$`52_T5_STM2457_01`$conversionRate)+as.numeric(S4U.data$`53_T5_STM2457_02`$conversionRate)+as.numeric(S4U.data$`54_T5_STM2457_03`$conversionRate)+as.numeric(S4U.data$`55_T5_STM2457_04`$conversionRate))/4

data_STM2457_T5$coverageOnTs_mean=(as.numeric(S4U.data$`52_T5_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`53_T5_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`54_T5_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`55_T5_STM2457_04`$coverageOnTs))/4

### STM2457 T6

data_STM2457_T6 <- S4U.data$`60_T6_STM2457_01`

data_STM2457_T6$conversionRate_mean=(as.numeric(S4U.data$`60_T6_STM2457_01`$conversionRate)+as.numeric(S4U.data$`61_T6_STM2457_02`$conversionRate)+as.numeric(S4U.data$`62_T6_STM2457_03`$conversionRate)+as.numeric(S4U.data$`63_T6_STM2457_04`$conversionRate))/4

data_STM2457_T6$coverageOnTs_mean=(as.numeric(S4U.data$`60_T6_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`61_T6_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`62_T6_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`63_T6_STM2457_04`$coverageOnTs))/4

### STM2457 T7

data_STM2457_T7 <- S4U.data$`68_T7_STM2457_01`

data_STM2457_T7$conversionRate_mean=(as.numeric(S4U.data$`68_T7_STM2457_01`$conversionRate)+as.numeric(S4U.data$`69_T7_STM2457_02`$conversionRate)+as.numeric(S4U.data$`70_T7_STM2457_03`$conversionRate)+as.numeric(S4U.data$`71_T7_STM2457_04`$conversionRate))/4

data_STM2457_T7$coverageOnTs_mean=(as.numeric(S4U.data$`68_T7_STM2457_01`$coverageOnTs)+as.numeric(S4U.data$`69_T7_STM2457_02`$coverageOnTs)+as.numeric(S4U.data$`70_T7_STM2457_03`$coverageOnTs)+as.numeric(S4U.data$`71_T7_STM2457_04`$coverageOnTs))/4


STM2457_4SU_mean_data <- list(data_STM2457_T0,
                           data_STM2457_T1,
                           data_STM2457_T2,
                           data_STM2457_T3,
                           data_STM2457_T4,
                           data_STM2457_T5,
                           data_STM2457_T6,
                           data_STM2457_T7)

names(STM2457_4SU_mean_data) <- c("STM2457_T0",
                           "STM2457_T1",
                           "STM2457_T2",
                           "STM2457_T3",
                           "STM2457_T4",
                           "STM2457_T5",
                           "STM2457_T6",
                           "STM2457_T7")
```

## Substract background T2C from 4SU data

In order to normalize T2C transitions to the background transition rate, conversion rates of the unlabelled samples are substracted from all conversion rates of labelled samples


### DMSO
```{r Background substraction, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}


# no 4SU data is >>>data_DMSO_no4SU 

##Substract background
for (i in 1:length(DMSO_4SU_mean_data)){
  DMSO_4SU_mean_data[[i]]$conversionRate_mean_background_sub <- DMSO_4SU_mean_data[[i]]$conversionRate_mean-data_DMSO_no4SU$conversionRate_mean}

# Normalize zu chase onset
onset_DMSO <- DMSO_4SU_mean_data$DMSO_T0

chase <- DMSO_4SU_mean_data


for (i in 1:length(chase)){
  chase[[i]]$norm_T2C <- chase[[i]]$conversionRate_mean_background_sub/onset_DMSO$conversionRate_mean_background_sub}

#give dset name column
for (i in 1:length(chase)){
  chase[[i]]$timepoints <- as.numeric(gsub("(DMSO_T)*", "", names(chase[i])))
}



for (i in 1:length(chase)){
  chase[[i]]$gene_coords <-paste0(chase[[i]]$Chromosome, chase[[i]]$Start, ":", chase[[i]]$End)
}


chase_finale<-do.call("rbind",chase)

chase_filtered=na.omit(chase_finale)

chase_filtered <- chase_filtered[!is.infinite(chase_filtered$norm_T2C),]





```


## STM2457
```{r Background substraction STM2457, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}




##Substract background
for (i in 1:length(STM2457_4SU_mean_data)){
  STM2457_4SU_mean_data[[i]]$conversionRate_mean_background_sub <- STM2457_4SU_mean_data[[i]]$conversionRate_mean-data_STM2457_no4SU$conversionRate_mean}

# Normalize zu chase onset
onset_STM2457 <- STM2457_4SU_mean_data$STM2457_T0

chase_STM <- STM2457_4SU_mean_data


for (i in 1:length(chase_STM)){
  chase_STM[[i]]$norm_T2C <- chase_STM[[i]]$conversionRate_mean_background_sub/onset_STM2457$conversionRate_mean_background_sub}

#give dset name column
for (i in 1:length(chase_STM)){
  chase_STM[[i]]$timepoints <- as.numeric(gsub("(STM2457_T)*", "", names(chase_STM[i])))
}






chase_finale_STM<-do.call("rbind",chase_STM)

chase_filtered_STM=na.omit(chase_finale_STM)

chase_filtered_STM <- chase_filtered_STM[!is.infinite(chase_filtered_STM$norm_T2C),]
#halflife
chase_filtered_STM$timepoints_experiment <- chase_filtered_STM$timepoints

chase_filtered_STM[chase_filtered_STM$timepoints==2,]$timepoints <- 1.5
chase_filtered_STM[chase_filtered_STM$timepoints==3,]$timepoints <- 2.3
chase_filtered_STM[chase_filtered_STM$timepoints==4,]$timepoints <- 3.5
chase_filtered_STM[chase_filtered_STM$timepoints==5,]$timepoints <- 5.3
chase_filtered_STM[chase_filtered_STM$timepoints==6,]$timepoints <- 8
chase_filtered_STM[chase_filtered_STM$timepoints==7,]$timepoints <- 12



```


# Expression distribution

We decided to use the coverage on Ts as an expression estimate. Cutoff >100 coverage on T. 


```{r Expression distribution, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}

#DMSO
ggplot(chase_filtered)+
  geom_density(aes(x=log10(coverageOnTs)))+
  geom_vline(xintercept = log10(100))+
  ggtitle("DMSO, cutoff 100")-> expression_DMSO



ggplot(chase_filtered_STM)+
  geom_density(aes(x=log10(coverageOnTs)))+
  geom_vline(xintercept = log10(100))+
  ggtitle("STM2457, cutoff 100")-> expression_stm



grid.arrange(expression_DMSO, expression_stm, ncol=2)








```




T coverage of 100: reads are length 84, after trimming ~70. In 70 bp, probably around 17 Ts are present. to have a coverage of 100 we would need only around 6 reads. 

T coverage of 1500: reads are length 84, after trimming ~70. In 70 bp, probably around 17 Ts are present. to have a coverage of 1500 we would need only around 90 reads. 

# Prepare halflife

+ Merge replicates, get mean C>T from foure replicates
+ Subtract T>C conversions of each time point from the T>C of No 4SU treatment (background subtraction)
+ Normalize to chase onset (Divide by chase onset C>T 0h)
+ Curve fitting using integrated rate law for a first-order reaction (min-pack.lm package) lineqr regression. 
+ Set half lifes >24 h to 24 h

```{r Halflife calculation DMSO, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}

calculate_model2 <- function(gene_data){
  tryCatch({
    model = nlsLM(norm_T2C ~ exp(-k * timepoints),
            data = gene_data,
            start=list(k = 0.5),
            lower = 0,
            control = nls.lm.control(maxiter = 1000),
            na.action = na.omit)
    return(broom::augment(model, gene_data) %>%
             mutate(halflife = log(2)/coef(model)["k"],
                    residual_sd_error = summary(model)$sigma))
  }, error = function(err){gene_data %>%
      mutate(.fitted = NA, .resid = NA, halflife = NA, residual_sd_error=NA)})
}

chase_filtered$timepoints_experiment <- chase_filtered$timepoints

chase_filtered[chase_filtered$timepoints==2,]$timepoints <- 1.5
chase_filtered[chase_filtered$timepoints==3,]$timepoints <- 2.3
chase_filtered[chase_filtered$timepoints==4,]$timepoints <- 3.5
chase_filtered[chase_filtered$timepoints==5,]$timepoints <- 5.3
chase_filtered[chase_filtered$timepoints==6,]$timepoints <- 8
chase_filtered[chase_filtered$timepoints==7,]$timepoints <- 12

  
result_TA_DMSO <- chase_filtered%>%
  dplyr::group_by(gene_name) %>%
  dplyr::mutate(T_coverage_mean = mean(coverageOnTs_mean))%>%
  dplyr::filter(T_coverage_mean>=100)%>%
  dplyr::group_modify(~ calculate_model2(.)) %>%
  dplyr::bind_rows()

result_TA_halflife_DMSO <- result_TA_DMSO %>%
  dplyr::select(gene_name, halflife, residual_sd_error,T_coverage_mean) %>%
  unique()

result_TA_halflife_DMSO <- as.data.frame(result_TA_halflife_DMSO)


result_TA_halflife_DMSO<- result_TA_halflife_DMSO[!is.infinite(result_TA_halflife_DMSO$halflife),]


# Set high halflifes to 24 

result_TA_halflife_DMSO[result_TA_halflife_DMSO$halflife>24,]$halflife <- 24

result_TA_halflife_DMSO$dset <- "DMSO"

#get gene anno

genes <- import.gff("~/Documents/Annotations/gencode.vM23.chr_patch_hapl_scaff.annotation.gtf", format="gtf", feature.type="gene")
genes<- genes[seqnames(genes)%in%level_order_mouse]
genes$chr <- as.character(seqnames(genes))

result_TA_halflife_DMSO$chr = NA
result_TA_halflife_DMSO$chr<-sapply(result_TA_halflife_DMSO$gene_name, function(x) unique(as.character(seqnames(genes[genes$gene_name==x]))))


result_TA_halflife_DMSO$chr <- as.character(result_TA_halflife_DMSO$chr)


result_TA_halflife_DMSO<-result_TA_halflife_DMSO[result_TA_halflife_DMSO$chr%in%level_order_mouse,]


result_TA_halflife_DMSO$gene_id = NA
result_TA_halflife_DMSO$gene_id<-sapply(result_TA_halflife_DMSO$gene_name, function(x) unique(as.character(genes[genes$gene_name==x][1,]$gene_id)))

result_TA_halflife_DMSO$gene_type = NA
result_TA_halflife_DMSO$gene_type<-sapply(result_TA_halflife_DMSO$gene_name, function(x) unique(as.character(genes[genes$gene_name==x][1,]$gene_type)))




result_TA_halflife_DMSO<- result_TA_halflife_DMSO[result_TA_halflife_DMSO$gene_type=="protein_coding",]



result_TA_halflife_DMSO<- result_TA_halflife_DMSO[result_TA_halflife_DMSO$chr%in%level_order_mouse,]

```


## Halflife STM
```{r Halflife calculation STM, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}


result_TA_STM2457 <- chase_filtered_STM %>%
  dplyr::group_by(gene_name) %>%
  dplyr::mutate(T_coverage_mean = mean(coverageOnTs_mean))%>%
  dplyr::filter(T_coverage_mean>=100)%>%
  dplyr::group_modify(~ calculate_model2(.)) %>%
  dplyr::bind_rows()


result_TA_halflife_STM2457 <- result_TA_STM2457 %>%
  dplyr::select(gene_name, halflife, residual_sd_error, T_coverage_mean) %>%
  unique()


result_TA_halflife_STM2457 <- as.data.frame(result_TA_halflife_STM2457)



result_TA_halflife_STM2457<- result_TA_halflife_STM2457[!is.infinite(result_TA_halflife_STM2457$halflife),]





result_TA_halflife_STM2457$chr = NA
result_TA_halflife_STM2457$chr<-sapply(result_TA_halflife_STM2457$gene_name, function(x) unique(as.character(seqnames(genes[genes$gene_name==x]))))


result_TA_halflife_STM2457$gene_id = NA
result_TA_halflife_STM2457$gene_id<-sapply(result_TA_halflife_STM2457$gene_name, function(x) unique(as.character(genes[genes$gene_name==x][1,]$gene_id)))

result_TA_halflife_STM2457$gene_type <- NA
result_TA_halflife_STM2457$gene_type<-sapply(result_TA_halflife_STM2457$gene_name, function(x) unique(as.character(genes[genes$gene_name==x][1,]$gene_type)))


result_TA_halflife_STM2457[result_TA_halflife_STM2457$halflife>24,]$halflife <- 24


result_TA_halflife_STM2457<- result_TA_halflife_STM2457[result_TA_halflife_STM2457$gene_type=="protein_coding",]




```


# Evaluate fitting of halflife
The residual standard error is the standard deviation of the residuals
– Smaller residual standard error means predictions are better

The Residual Standard Error is the average amount that the response (dist) will deviate from the true regression line.

Residual standard error =

where:

y: The observed value
yi: The predicted value
df: The degrees of freedom, calculated as the total number of observations – total number of model parameters.

The residuals of this model (the difference between the observed values and the predicted values) will be small, which means the residual standard error will also be small.

```{r Evaluating, echo=FALSE, eval=TRUE, warning=FALSE,message=FALSE}

ggplot(data=result_TA_halflife_DMSO)+
  geom_density(aes(x=log10(residual_sd_error)))


ggplot(data=result_TA_halflife_DMSO)+
  stat_ecdf(aes(x=residual_sd_error))+
  coord_cartesian(xlim=c(0,5))+
  geom_vline(xintercept = 0.3, linetype="dotted")


paste0("Number of UTRs unfiltered:", nrow(result_TA_halflife_DMSO))

result_TA_halflife_DMSO_filtered <- result_TA_halflife_DMSO[result_TA_halflife_DMSO$residual_sd_error<0.3,]

paste0("Number of UTRs unfiltered:", nrow(result_TA_halflife_DMSO_filtered))



```


We decided to do the residual error cutoff at 0.3.




# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
